\LARGE \bf Multi-Label Speech Emotion Recognition Using 2D Convolutional Neural Networks

### 2019-10-11 research/paper FG/paper.tex
Update paper
Update paper and submit to FG2020

%use this in case of several affiliations
%\author{\parbox{16cm}{\centering
%    {\large Huibert Kwakernaak$^1$ and Pradeep Misra$^2$}\\
%    {\normalsize
%    $^1$ Faculty of Electrical Engineering, Mathematics and Computer Science, University of Twente, Enschede, The Netherlands\\
%    $^2$ Department of Electrical Engineering, Wright State University, Dayton, USA}}
%    \thanks{This work was not supported by any organization}% <-this % stops a space
%}

@@

The main idea of this paper is to build a more realistic SER system that can recognize multiple emotions in speech. We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions. We used four public databases then combined the databases into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers. The final output of the neural network is a list of present emotions.

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on the previous work by \cite{Engel2019} and \cite{Badshah2017} where log-Mel spectrograms were successfully used for extracting audio features.

The choice to use a CNN is based on the previous work of Balakrishnan et al. \cite{Balakrishnan2017} which found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due to treating the log-Mel spectrograms as images and CNNs have been shown to perform well on images \cite{Krizhevsky2012}. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} libraries respectively.

@@

%\begin{table*}
%	\centering
%	\caption{Comparison of databases with their labeled emotions.}
%	\label{dbEmoTable}
%	\begin{tabular}{@{}lccccccccccccc@{}}
%		\toprule
%		Databases & \multicolumn{13}{c}{Emotions} \\ 
%		\cmidrule{2-14}
%		& Neutral & Anger & Disgust & Fear & Happy & Sad & Surprise & Calm & Excitement & Frustration & Amused & Sleepy & Bored \\ 
%		\midrule
%		IEMOCAP & x & x & x & x & x & x & x &   & x & x &   &   &   \\
%		CREMA-D & x & x & x & x & x & x &   &   &   &   &   &   &   \\
%		TESS & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		RAVDESS & x & x & x & x & x & x & x & x &   &   &   &   &   \\
%		MSP-IMPROV & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		SAVEE & x & x & x & x & x & x & x &   &   &   &   &   &   \\
%		Emo-DB & x & x & x & x & x & x &   &   &   &   &   &   & x \\
%		EmoV-DB & x & x & x &   &   &   &   &   &   &   & x & x &   \\
%		\bottomrule
%	\end{tabular}%
%\end{table*}
%Table \ref{dbEmoTable} compares the set of emotions of each database which makes it easier to identify overlapping emotions.

@@

To control for the audio quality, we resampled all samples to 48 kHz, applied noise reduction to samples from the IEMOCAP and CREMA-D databases, and cropped all samples to 4.5 seconds. We used a kaiser filter for resampling and used spectral gating for noise reduction. We only applied noise reduction to the IEMOCAP and CREMA-D databases after listening to the audio. Lastly, we chose databases that are spoken in English to maintain language consistency.

@@

%\begin{figure}
%	\centering
%	\hspace{6mm}
%	\includegraphics[width=\linewidth]{high_level_dataflow_diagram.png}
%	\caption{High level overview of the processing stages that a speech sample goes through.}
%	\label{highLevelDataFlowDiagram}
%\end{figure}

@@

%\begin{figure*}
%	\centering
%	\hspace{6mm}
%	\includegraphics[width=\textwidth]{neural_network_architecture.png}
%	\caption{Architecture for the SER network including the input layer and eight processing layers.}
%	\label{neuralNetworkArchitecture}
%\end{figure*}

@@

Table \ref{modelHyperparams} describes the model's hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters. The decision to use \textit{binary cross-entropy} as the loss measure is due to multi-label problems considering each output as independent and thus each output acts like a binary classifier. This lets the model only penalize the outputs that are incorrect without penalizing the other outputs. However, we use \textit{categorical cross-entropy} as the accuracy metric because binary cross-entropy would provide the incorrect accuracy as it defined for a binary output such as 0 or 1, not vectors containing binary outputs. We could have picked a different accuracy metric but chose categorical cross-entropy due to its ease of use.

>>>>

@@

The main idea of this paper is to build a more realistic SER system that can recognize multiple emotions in speech. We approach this problem by using more data and by using a 2D CNN to classify speech samples into multiple emotions. We used four public databases and combined them into an a single database by processing all of the speech samples into log-Mel spectrograms. These log-Mel spectrograms were then fed into an eight-layer neural network consisting of four convolutional layers and four dense layers. The final output of the neural network is a list of the emotions present in the speech sample.

@@

We approach the problem of SER by preprocessing samples from four databases and then feeding those samples into a CNN. We preprocess the speech samples into log-Mel spectrogram to help the neural network extract features relevant to emotion recognition. This choice is based on the work by \cite{Engel2019} and \cite{Badshah2017} where log-Mel spectrograms were successfully used for extracting audio features.

The choice to use a CNN is based on the work of Balakrishnan et al. \cite{Balakrishnan2017} which found that CNN-based models have superior performance compared to RNNs. Another justification for using CNNs is due to treating the log-Mel spectrograms as images and CNNs have been shown to perform well on images \cite{Krizhevsky2012}. Both the preprocessing steps and the CNN were implemented in Python using the Librosa \cite{McFee2015} and Keras \cite{Chollet2015} libraries respectively.

@@

@@

To control for the audio quality, we resampled all samples to 48 kHz, applied noise reduction to samples from the IEMOCAP and CREMA-D databases, and cropped all samples to 4.5 seconds. We used a kaiser filter for resampling and we used spectral gating for noise reduction. We only applied noise reduction to the IEMOCAP and CREMA-D databases after listening to the audio. Lastly, we chose databases that are spoken in English to maintain language consistency.

@@

@@

@@

Table \ref{modelHyperparams} describes the model's hyperparameters and table \ref{trainHyperparams} describes the training hyperparameters. The decision to use \textit{binary cross-entropy} as the loss measure is due to multi-label problems considering each output as independent and thus each output acts like a binary classifier. This lets the model only penalize the outputs that are incorrect without penalizing the other outputs. However, we use \textit{categorical cross-entropy} as the accuracy metric because binary cross-entropy would provide the incorrect accuracy as it defined for a binary output such as 0 or 1, not vectors containing binary outputs. Different loss functions should be considered in the future to improve training results.
